---
title: "Installation and troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation and troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Requirements

polyglotSQL has a compiled Rust backend. To install **from source** you need:

* R >= 4.2
* Rust: **cargo** and **rustc >= 1.88**
* the `xz` utility (to unpack the vendored dependencies; present on
  virtually every system)

No Rust is needed when installing a prebuilt binary (e.g. from CI artifacts
or a binary repository).

## Installing Rust

The recommended path on every platform is [rustup](https://rustup.rs):

```sh
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh
```

Alternatives: `brew install rust` (macOS), `apt-get install cargo rustc`
(Debian/Ubuntu — check the version!), the official Windows installer from
<https://forge.rust-lang.org/infra/other-installation-methods.html>. On
Windows, the `gnu` toolchain targeting `x86_64-pc-windows-gnu` is required
for R; rustup installs it with:

```sh
rustup target add x86_64-pc-windows-gnu
```

The package's `configure` script checks the toolchain up front and prints
exactly what is missing (it never installs anything itself).

## Installing the package

```{r}
# install.packages("remotes")
remotes::install_github("milkway/polyglot-sql-r")
```

## Offline / air-gapped builds

The source package ships all Rust dependencies as a vendored archive
(`src/rust/vendor.tar.xz`), and the build runs `cargo` with `--offline`.
**Nothing is downloaded during `R CMD INSTALL`** — an internet connection is
only needed to obtain the package itself.

## Build details you may care about

* Compilation uses at most **2 parallel jobs** (CRAN policy).
* The Rust static library is linked into a single shared object; the
  installed package has no runtime dependency on cargo.
* `DEBUG=true R CMD INSTALL .` produces a debug build (faster compile,
  slower runtime).

## Troubleshooting

### `cargo: command not found` / configure fails

Rust is not on the `PATH` R uses. After installing via rustup, restart R so
`~/.cargo/bin` (added to your shell profile) is visible, or add it manually:

```r
Sys.setenv(PATH = paste0(Sys.getenv("PATH"), ":", path.expand("~/.cargo/bin")))
```

### `rustc >= 1.88 required`

Distribution packages are often too old. Install via rustup, or update with
`rustup update stable`.

### Linker errors on macOS

Install the Xcode command-line tools: `xcode-select --install`.

### Windows: `cannot find -lpolyglotsql` or wrong target

Make sure the `x86_64-pc-windows-gnu` target is installed (see above) and
that you are using Rtools matching your R version.

### Updating the embedded engine (for developers)

The upstream crate version is pinned in `src/rust/Cargo.toml` and
`Cargo.lock`, and its sources are vendored. To upgrade:

```sh
tools/update-vendor.sh 0.6.3   # bumps polyglot-sql, re-vendors, re-packs
```

then update `Config/polyglotSQL/upstream` in `DESCRIPTION`, run the test
suite, and check `polyglot_version()` reports the new version.

## Verifying an installation

```{r}
library(polyglotSQL)
polyglot_version()
sql_transpile("SELECT IFNULL(a, b) FROM t", from = "mysql", to = "postgres")
```
